Joint Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2024-03-12 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn]{aastex631}

% Filler text
\usepackage{blindtext}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage[autostyle]{csquotes}
\usepackage{amssymb}
\usepackage{makecell}
\let\tablenum\relaxgit 
\usepackage{siunitx}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\arcsinh}{arcsinh}

%\renewcommand\theadalign{bc}

@@

\newcommand{\vlk}[1]{{\color{blue} [VLK: #1]}}
\newcommand{\as}[1]{{\color{purple} [AS: #1]}}
\newcommand{\dvd}[1]{{\color{red} [DvD: #1]}}
% Begin!
\begin{document}

% Title
    \title{Joint Deconvolution of Astronomical Images in the Presence of Poisson Noise}
    %Jolideco: A Simultaneous Deconvolution of Multiple Images in the Presence of Poisson Noise}

@@

% Abstract with filler text
    \begin{abstract}

        We present a new method for joint likelihood deconvolution (\jolideconospace) of a set of astronomical observations of the same sky region in the presence of Poisson noise. 
        The observations may be obtained from different instruments with  different resolution, and different point spread functions. \jolideco reconstructs a single flux image by optimizing the posterior distribution based on the joint Poisson likelihood of all observations under a patch-based image prior. The patch prior is parameterised via a Gaussian Mixture model which we train on high-signal-to-noise astronomical images, including data from the  James Webb Telescope and the GLEAM radio survey. This prior favors correlation structures among the reconstructed pixel intensities that are characteristic of those observed in the %astronomical 
        training images. 
        % used to train the prior. 
        It is, however, not informative for the mean or scale of the reconstruction. By applying the method to simulated data we show that the combination of multiple observations and the patch-based prior leads to much improved reconstruction quality in many different source scenarios and signal to noise regimes. We demonstrate that \jolideco yields superior reconstruction quality relative to alternative standard methods such as the Richardson-Lucy method. We illustrate the results of \jolideco\ applied to example data from the \chandra observatory and the \ferminospace. By comparing the measured width of a counts based and the corresponding \jolideco flux profile we find the deconvolved width to be consistent with the theoretical expectation derived from the known width of the PSF.
    \end{abstract}

    % Main body with filler text

    \section{Introduction}
    The resolution of any physical or astronomical imaging process is inherently limited by the instrument or telescope used. Furthermore, the quality of the resulting image is affected by background, instrumental measurement noise, and non-uniform exposure. This is especially true in the low signal-to-noise regime, where images are affected by Poisson noise. This includes images taken by \xray and \gammaray~telescopes but also medical images obtained with methods such as positron emission tomography or single-photon emission computed tomography. This low counts imaging process is illustrated in Figure~\ref{fig:model-illustration}.

@@

    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/model-illustration.pdf}
            \caption{
                Illustration of the forward model of the low counts imaging process. The flux represents the true unknown flux distribution in the sky. The flux is convolved with an image of the PSF, which represents an estimate of the resolution of the imaging instrument. The exposure consists of the superposition of multiple independent observations with different observation time and pointing position. The exposure weighted flux image is degraded by Poisson noise.
            }
            \label{fig:model-illustration}
            \script{model-illustration.py}
        \end{centering}
    \end{figure*}

%\as{You may want to add a short para or just a sentance saying that X-ray and gamma-ray images are the Poisson type images containing low number of counts in each image pixel etc.. - this is to link to the next subsection which is specific to Poisson noise?}
% AD: this is mentioned already...

@@

    Other interesting approaches to the problem of deconvolution can of course be found in the field of computer vision. For example \cite{Zoran2011} build on very similar assumption of local correlation for natural images. However, instead of assuming a fixed local correlation structure, such as smoothness defined by a Gaussian kernel with a given bandwidth determining the correlation scale,
    %with a correlation size, 
    they proposed to learn the local correlation structure on millions of patches extracted from a set of reference images. For this they propose to split training images from representative data sets into small patches and learn a  Gaussian Mixture Model (GMM) %to model 
    for the distribution of extracted patches. To use the GMM in reconstruction they introduce the concept of expected patch likelihood (EPLL). They show that the patch-based GMM prior leads to much improved image reconstructions for multiple inverse problems such as denoising, deblurring and inpainting. \cite{Bouman2016} later adapted the patch prior successfully for the reconstruction of sub-millimeter radio data from the Event Horizon Telescope (EHT).

    % \begin{figure*}
    %     \script{rl_decomposition.py}
    %     \begin{centering}
    %         \includegraphics[width=\textwidth]{figures/richardson-lucy-decomposition.pdf}
    %         \caption{
    %             Decomposition of the RL algorithm for various number of iterations. \todo{Improve figure}
    %         }
    %         \label{fig:rl_decomposition}
    %     \end{centering}
    % \end{figure*}

@@

    \subsection{Poisson Joint Likelihood}
    Our goal is to recover an image $\mathbf{x}$ from a set of multiple low counts observations. Most generally we assume our total dataset consists of $J$ individual observations of the same region of the sky, which we jointly model. The assumption is that the underlying, unknown \enquote{true emission} image, we are interested in, does not change over the timescale of the observations. Under this assumption the datasets can be for example:

@@

    Using Bayes' rule we can express the \aposteriori likelihood under a given prior distribution, $P(\mathbf{x} )$:
    %
    \begin{equation}
        \label{eq:bayes}
        P(\mathbf{x}|\textbf{D}) = P(\mathbf{x} ) \frac{P(\textbf{D} |\mathbf{x})}{P(\textbf{D})}.
    \end{equation}
    %
    The Maximum A Posteriori (MAP) estimate of $\mathbf{x}$ is the value that maximizes Equation~\ref{eq:bayes}.
    Taking the logarithm of Equation~\ref{eq:bayes}, replacing the definition of the likelihood $P(\mathbf{D}|\mathbf{x})$ with Equation~\ref{eq:joint} and dropping the normalisation term $P(\mathbf{D})$ which is independent of $\mathbf{x}$, we obtain the following expression for the log-posterior distribution
    $\mathcal{L}$:
    %
    \begin{equation}
        \label{eq:total}
        \mathcal{L}\left(\mathbf{x} | \mathbf{D} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D}_j | \mathbf{x} \right) - \beta \cdot \mathcal{P}(\mathbf{x}),
    \end{equation}
    %
    where $\mathcal{C}\left( \mathbf{D}_j | \mathbf{x} \right)$ represents the summed log-likelihood for an individual observation $j$ and $\mathcal{P}(\mathbf{x})$ represents the log of the prior distribution with the factor $\beta$ adjusting the weight of the prior relative to the joint log likelihood term. The parameter $\beta$ can be viewed as a hyper-parameter, chosen by the user to adjust the strength of the prior distribution relative to the likelihood. The choice of $\beta$ is optional and for all experiments presented in this paper we set $\beta=1$.

    %\newpage
    \subsection{patch-based Image Prior}
    \label{sec:patch-prior}
    In general it is difficult to capture global image characteristics in a prior distribution taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the squre of the number of pixels  and thus quickly become computationally untraceable. However, astronomical images typically only exhibit local correlations among the pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light. This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the patch-based image prior distribution introduced by \cite{Zoran2011}.

    A patch is defined as a small, typically square, region of size $N_p \times N_p$ pixels, representing a local neighbourhood in a given image. Mathematically we can formulate the extraction of such patches as:
    %
    \begin{equation}
        \mathbf{x}_n = \mathbf{P}_n \mathbf{x}
    \end{equation}
    %
    Where $\mathbf{P}_n$ represents a neighbourhood matrix to extract the $n$-th patch from the image $\mathbf{x}$ and where $\mathbf{x}_n$ is the vector of length $N_p^2$ representing the \enquote{flattened} image patch. By obtaining a large number of patches we can model the distribution of pixels values within the patches using a Gaussian Mixture Model (GMM). The GMM is a universal function approximator and thus can model arbitrary distributions. For now we assume that we have a pre-trained GMM which encodes the patch distribution of some reference image we have chosen. During the actual image reconstruction we aim to identify the most likely GMM component per patch, treating this as a discrete hyper-parameter
    that is optimized along with the image $\mathbf{x}$.

@@

    If we knew the relevant component, $k_n$, for patch $\mathbf{x}_n$ we could derive the conditional log-prior distribution directly from Equation~\ref{eq:gmm}:
    %
    \begin{equation}
    \begin{split}
    \mathcal{P}_{\rm GMM}(\mathbf{x}_n|k_n) = -\frac{1}{2} \left[ \right. d \log(2\pi)
    + \log(|\boldsymbol \Sigma_{k_n}|)\\
    + (\mathbf{x}_n - \boldsymbol{\mu}_{k_n})^T \Sigma^{-1}_{k}(\mathbf{x}_n - \boldsymbol{\mu}_{k_n}) \left. \right].
    \label{eq:cond.log.prior}
    \end{split}
    \end{equation}
    %
    If we split the image $\mathbf{x}$ into multiple non-overlapping patches, for each patch $\mathbf{x}_n$, we can find the GMM component $\hat{k}_n$, which maximises the conditional log-prior distribution given in Equation~\ref{eq:cond.log.prior}:
    %
    \begin{equation}
        \hat{k}_n = \underset{k \in \{1, ..., K\} }{\argmax}{\mathcal{P}_{\rm GMM}(\mathbf{x}_n|k)}
        \label{eq:khat}
    \end{equation}
    \vspace{0.2em}
    %
    As the number of components in the GMM is typically limited to $\mathcal{O}(10^2)$ we can optimize the conditional log-prior distribution by simply evaluating it for each of the GMM components and choose the one with the maximum value. Finally, to evaluate the (optimizied) log-prior of the full image we sum the maximum values of the individual patches: 
    %
    \begin{equation}
    \label{eq:prior-total}
        \mathcal{P}(\mathbf{x}, \hat{\mathbf{k}}) = 
        %\mathcal{P}(\mathbf{x} | \hat{\mathbf{k}})  + \mathcal{P}(\hat{\mathbf{k}}) =
        \sum_{n = 1}^N \mathcal{P}_{\rm GMM}(\mathbf{P}_n \mathbf{x} | \hat{k}_n) +
        \sum_{n = 1}^N \log \pi_{\hat{k}_n},
    \end{equation}   
    %
    where $\mathbf{P}_n$ is again the matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed.

@@

    Apart from this, the process for choosing the most likely GMM prior component remains unmodified. However, we note that we neglected the prior dependence between neighbouring patches\footnote{That is, we continue to evaluate the log-prior distribution of the full image by summing over the overlapping patches as in Equation~\ref{eq:prior-total}, even though overlapping patches include common pixes and are thus clearly not \aprioir independent.}. As the choice of the GMM component per patch is discrete, a simultaneous gradient based optimization of all patches is not easily possible, because we cannot compute gradients across GMM components. This would require a continuous representation or interpolation between GMM components.

    % \begin{itemize}
    %     \item the case of non-overlapping patches is rather simple
    %     \item for overlapping patches we have to make a few more approximations
    %     \item either we neglect the correlation between overlapping patches
    %     \item or we have to optimize the array of $\mathbf{k}$ at the same time, instead of per patch optimization. However optimization on discrete 
    %     values is not easily possible in Pytorch. However here is one idea:
    %     group the GMM components by similarity and introduce a single continues
    %     parameter $z$, which interpolates between similar components.
    %     \item Another idea is to use Gumbel Softmax...
    %     \item we also compute the mean of the likelihood for the overlapping parts
    % \end{itemize}

    % To use the learned GMM as prior for the reconstruction of another image, the image is split into overlapping patches. For each of the patches the log-likelihood for each of the GMM components is evaluated. The resulting grid of overlapping patches is illustrated in Figure~\ref{fig:patches}.

    % with radio astronomy data. They have shown that the reconstructed image only weakly depends on the choice of the reference data on which the the patch prior is learned. They found equivalent results for GMMs learned on natural images and specifically simulated images of black hole ring structures.

    % Using half quadratic splitting and the Wiener filter solution, which is not possible in our case. Instead we choose a direct optimization by relying on differentiable programming later.

    \subsubsection{Image Dynamic Range and Patch Normalisation}
    \label{ssec:patch-norm}
    Astronomical images typically show a much higher dynamic range compared to natural images.
    % This is mostly due to the existence of point sources, which rarely occur in natural images. Astronomical point sources typically relate to compact bright objects, such as unresolved active galactic nuclei at high distances, or binary objects and stars at Galactic distances. In addition the visible matter in the universe is sparse such that astronomical images typically show sparse bright emission on a dark background. 
    To normalise the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to be present at many levels of intensity. While this could be handled by introducing significantly more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we instead propose a modified normalisation for the patches:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \tilde{\mathbf{x}}_n = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}
    \end{equation}
    %
    By dividing by the mean of the patch, $\Bar{\mathbf{x}}_n$, the structure in the patch becomes independent of the total brightness the patch contains. This results in patches being grouped only by their pixel to pixel contrast and not by their overall brightness. This corresponds to the assumption that the structure in the patch is \aprioir independent of the total flux contained in the patch, which we consider a valid assumption for astronomical images. This also generalizes to the prior distribution of the whole image described by Equation~\ref{eq:prior-total}. The value of the prior distribution is independent of the total flux of the image. The prior distribution simply regularizes the structure in the image, while the overall intensities are typically well constrained by data as quantified by the Poisson likelihood term. Thus we simply assume a uniform prior distribution for the patch intensities. Algorithmically, we replace $\mathbf{x}_n$ with $\tilde{\mathbf{x}}_n$ when evaluating Equations~\ref{eq:gmm} \ref{eq:khat} and transform $\mathbf{P}_n\mathbf{x}$ via Equation~\ref{eq:patch-norm} when evaluating Equation~\ref{eq:prior-total}.

    % The high dynamic range of astronomical images challenges any deconvolution method. To recover point sources accurately it is required to keep correlations with neighbouring pixels low, while for extended source the opposite is true. Ideally those contradicting requirements can be handled with a single flexible prior assumption, which adapts to the data, such as the patch-based GMM prior introduced in the previous section.

    % With a given non-linearity:

    % \begin{equation}
    %     f(x) = \arcsinh{x / \alpha}
    % \end{equation}

    % \todo{The patch normalisation remains to be decided, but it would be advantageous to get rid of the hyper-parameter associated with the total image normalisation}.
    % The GMM patch prior is learned from other images whose pixel intensities
    % have no physical relationship of the intensities of the Poisson data we are
    % dealing with. To overcome this the GMM is learned on normalized images,
    % where the intensity values are constrained between 0 and 1.
    % To evaluate the GMM patch prior on the 
    % model image $\mathbf{x}$ the pixel values of $\mathbf{x}$  need to be mapped 
    % in the same dynamic range between 0 and 1. Multiple common choices for such a
    % mapping function are shown in Fig.~\ref{fig:image-norms}.

    % In practice we are interested in the likelihood of a given structure in a patch and
    % So we subtract the mean value of a given patch. This represents the assumption that
    % the image structure is independent of the total flux contained in the patch.

    % % \begin{figure}
    % %     \script{image-norms.py}
    % %     \begin{centering}
    % %         \includegraphics[width=\linewidth]{figures/image-norms.pdf}
    % %         \caption{
    % %             Various image norms with different dynamic behaviour.
    % %         }
    % %         \label{fig:image-norms}
    % %     \end{centering}
    % % \end{figure}

    % The choice of the image normalisation allows to adjust the contrast of the image
    % and for example enhance low intensity structures, which also enhances the regularising
    % effect of the patch prior.

    % \todo{Which normalisation to choose, atan, asinh, log, inverse-cdf?}

    % There are arguments for $\arcsinh(\mathbf{x})$, see for example \cite{Lupton2004}, which we use.

@@

    % \subsection{Choice and Tuning of Hyperparameters}
    % \subsubsection{Cross Validation}
    % Cross validation (CV) is an established method for model selection in statistics and machine learning to prevent a model from overfitting to the training data. To perform CV the available data is slit into a \enquote{training} and \enquote{validation} dataset. While optimizing the log-likelihood value is monitored  The use of multiple dataset allows to use cross-validation to prevent overfitting. One of the standard techniques in machine learning...

    % Cross validation is feasible once sufficient data is available, which is for example the 
    % case for deep \chandra~observations.

    % CV as stopping rule for RL was proposed by \cite{Reeves1995}...

    % \subsubsection{Other Approaches}

    % Self supervised Poisson loss? J-invariance has been shown to work, Grid Search? Calibration on similar data with ground truth?
    % Ground truth based calibration: record prior vs data likelihood on simulated data...

    \subsection{Systematic Errors of Predicted Counts}
    Each observation $\mathbf{D}_j$ is subject to systematic errors. For example, the pointing accuracy of the telescope when repeatedly pointed to the same object or when using different types of telescopes for observation, the absolute position of an astronomical object may vary among the  observations. The process of correcting for these inaccuracies is called \textit{astrometry} and is a standard procedure in for example \chandra data analysis. A second example of systematic errors is the variation of background emission between multiple observations.

    To account for both systematic effects we extend the model in Equation~\ref{eq:model} for the predicted counts, by a background normalisation factor $\alpha_j$ and a linear shift in position $\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})$, which is specific for each observation. This yields the following modified model:
    %
    \begin{equation}
        \label{eq:model-npred-calibration}
        \mathbf{\lambda}_j = \mathbf{E}_j \cdot  \left(\mathrm{PSF}_j \circledast \phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})\right) + \alpha_j \cdot \mathbf{B}_j
    \end{equation}
    %
    In practice one observation is used as a reference and its corresponding systematic shifts in position $\delta_{x,j}, \delta_{y,j}$ are frozen during optimization. An example of the effect of the positional shift and background normalisation can be seen in Figure~\ref{fig:example-fermi-lat-non-calibrated}.

@@

    We also experimented with computing the diagonal of the Hessian of the posterior distribution evaluated at the MAP estimate. However we found this numerically challenging because of random the cycle spinning.

@@

    We implemented the \jolideco method as an independent Python package, based on PyTorch as the computational back-end.  We defined a modular, object oriented code structure, to allow parts of the algorithm to be flexible and interchangeable, such as the choice of the patch normalisation, the choice of GMM, optimization methods, and serialisation formats for the reconstruction results as well as the corresponding diagnostic information, such as the trace of the posterior and prior likelihood values. This will simplify future improvement and extension of the package as well as the adoption of \jolideco by the astronomical community. The package is available on the platform GitHub\footnote{\url{https://github.com/jolideco/jolideco}}.

@@

    \subsection{Learning GMMs}
    \label{ssec:jolideco-gmm-library}
    The patch-based image prior described in Section~\ref{sec:patch-prior} requires a pre-trained GMM. We train the GMM by choosing a set of reference images, splitting them into patches of size $8\times8$ pixels and learning a GMM with $K$ components from the whole distribution of patches. In general the strength of the GMM prior is its high \enquote{expressiveness}. This means the prior distribution contains a large variety of structural shapes and has freedom to adapt to the data in the optimization process by choosing the most likely GMM component for each patch in the image. With radio data, \cite{Bouman2016} found their results to be mostly independent of the choice of data the GMM prior was learned on.

@@

    % To evaluate the reconstruction performance of \jolideco and compare it against other methods we defined a set of test scenarios. They consist of simplified elementary morphological structures resembling those found in astronomical images. We have devised four distinct arrangements of combinations of point and extended sources of different shapes. The scenarios are:

>>>>

% Define document class
\documentclass[twocolumn, lineno]{aastex631}

\usepackage{blindtext}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{showyourwork}
\usepackage[autostyle]{csquotes}
\usepackage{amssymb}
\usepackage{makecell}
\let\tablenum\relaxgit 
\usepackage{siunitx}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\arcsinh}{arcsinh}

@@

\newcommand{\vlk}[1]{{\color{blue} [VLK: #1]}}
\newcommand{\as}[1]{{\color{purple} [AS: #1]}}
\newcommand{\dvd}[1]{{\color{red} [DvD: #1]}}

\begin{document}

% Title
    \title{Joint Deconvolution of Astronomical Images in the Presence of Poisson Noise}

@@

    \begin{abstract}

        We present a new method for joint likelihood deconvolution (\jolideconospace) of a set of astronomical observations of the same sky region in the presence of Poisson noise. 
        The observations may be obtained from different instruments with  different resolution, and different point spread functions. \jolideco reconstructs a single flux image by optimizing the posterior distribution based on the joint Poisson likelihood of all observations under a patch-based image prior. The patch prior is parameterised via a Gaussian Mixture model which we train on high-signal-to-noise astronomical images, including data from the  James Webb Telescope and the GLEAM radio survey. This prior favors correlation structures among the reconstructed pixel intensities that are characteristic of those observed in the training images. It is, however, not informative for the mean or scale of the reconstruction. By applying the method to simulated data we show that the combination of multiple observations and the patch-based prior leads to much improved reconstruction quality in many different source scenarios and signal to noise regimes. We demonstrate that with the patch prior \jolideco yields superior reconstruction quality relative to alternative standard methods such as the Richardson-Lucy method. We illustrate the results of \jolideco\ applied to example data from the \chandra X-ray Observatory and the \ferminospace\ Gamma-ray Space Telescope. By comparing the measured width of a counts based and the corresponding \jolideco flux profile of an X-ray filament in SNR {\it 1E 0102.2-7219} we find the deconvolved width of $0.58\pm 0.02$ arcsec to be consistent with the theoretical expectation derived from the known width of the PSF.
    \end{abstract}

    \section{Introduction}
    The resolution of any physical or astronomical imaging process is inherently limited by the instrument or telescope used. Furthermore, the quality of the resulting image is affected by background, instrumental measurement noise, and non-uniform exposure. This is especially true in the low signal-to-noise regime, where images are affected by Poisson noise. This includes images taken by \xray and \gammaray~telescopes but also medical images obtained with methods such as positron emission tomography or single-photon emission computed tomography. This low-count imaging process is illustrated in Figure~\ref{fig:model-illustration}.

@@

    \begin{figure*}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/model-illustration.pdf}
            \caption{
                Illustration of the forward model of the low-count imaging process. The flux represents the true unknown flux distribution in the sky. The flux is convolved with an image of the PSF, which represents an estimate of the resolution of the imaging instrument. The exposure consists of the superposition of multiple independent observations with different observation time and pointing position. The exposure weighted flux image is degraded by Poisson noise.
            }
            \label{fig:model-illustration}
            \script{model-illustration.py}
        \end{centering}
    \end{figure*}

@@

    Other interesting approaches to the problem of deconvolution can of course be found in the field of computer vision. For example \cite{Zoran2011} build on very similar assumption of local correlation for natural images. However, instead of assuming a fixed local correlation structure, such as smoothness defined by a Gaussian kernel with a given bandwidth determining the correlation scale, they proposed to learn the local correlation structure on millions of patches extracted from a set of reference images. For this they propose to split training images from representative datasets into small patches and learn a  Gaussian Mixture Model (GMM) for the distribution of extracted patches. To use the GMM in reconstruction they introduce the concept of expected patch likelihood (EPLL). They show that the patch-based GMM prior leads to much improved image reconstructions for multiple inverse problems such as denoising, deblurring and inpainting. \cite{Bouman2016} later adapted the patch prior successfully for the reconstruction of sub-millimeter radio data from the Event Horizon Telescope (EHT).

@@

    \subsection{Poisson Joint Likelihood}
    Our goal is to recover an image $\mathbf{x}$ from a set of multiple low-count observations. Most generally we assume our total dataset consists of $J$ individual observations of the same region of the sky, which we jointly model. The assumption is that the underlying, unknown \enquote{true emission} image, we are interested in, does not change over the timescale of the observations. Under this assumption the datasets can be for example:

@@

    Using Bayes' rule we can express the \aposteriori likelihood under a given prior distribution, $P(\mathbf{x} )$:
    %
    \begin{equation}
        \label{eq:bayes}
        P(\mathbf{x}|\textbf{D}) = P(\mathbf{x} ) \frac{P(\textbf{D} |\mathbf{x})}{P(\textbf{D})}.
    \end{equation}
    %
    The Maximum A Posteriori (MAP) estimate of $\mathbf{x}$ is the value that maximizes Equation~\ref{eq:bayes}.
    Taking the logarithm of Equation~\ref{eq:bayes}, replacing the definition of the likelihood $P(\mathbf{D}|\mathbf{x})$ with Equation~\ref{eq:joint} and dropping the normalization term $P(\mathbf{D})$ which is independent of $\mathbf{x}$, we obtain the following expression for the log-posterior distribution
    $\mathcal{L}$:
    %
    \begin{equation}
        \label{eq:total}
        \mathcal{L}\left(\mathbf{x} | \mathbf{D} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D}_j | \mathbf{x} \right) - \beta \cdot \mathcal{P}(\mathbf{x}),
    \end{equation}
    %
    where $\mathcal{C}\left( \mathbf{D}_j | \mathbf{x} \right)$ represents the summed log-likelihood for an individual observation $j$ and $\mathcal{P}(\mathbf{x})$ represents the log of the prior distribution with the factor $\beta$ adjusting the weight of the prior relative to the joint log likelihood term. The parameter $\beta$ can be viewed as a hyper-parameter, chosen by the user to adjust the strength of the prior distribution relative to the likelihood. The choice of $\beta$ is optional and for all experiments presented in this paper we set $\beta=1$.

    %\newpage
    \subsection{Patch-based Image Prior}
    \label{sec:patch-prior}
    In general it is difficult to capture global image characteristics in a prior distribution taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the squre of the number of pixels  and thus quickly become computationally untraceable. However, astronomical images typically only exhibit local correlations among the pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light. This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the patch-based image prior distribution introduced by \cite{Zoran2011}.

    A patch is defined as a small, typically square, region of size $N_p \times N_p$ pixels, representing a local neighbourhood in a given image. Mathematically we can formulate the extraction of such patches as:
    %
    \begin{equation}
        \mathbf{x}_n = \mathbf{P}_n \mathbf{x},
    \end{equation}
    %
    where $\mathbf{P}_n$ represents a neighbourhood matrix to extract the $n$-th patch from the image $\mathbf{x}$ and where $\mathbf{x}_n$ is the vector of length $N_p^2$ representing the \enquote{flattened} image patch. By obtaining a large number of patches we can model the distribution of pixel intensities within the patches using a Gaussian Mixture Model (GMM). The GMM is a universal function approximator and thus can model arbitrary distributions. For now we assume that we have a pre-trained GMM which encodes the patch distribution of some reference image we have chosen. During the actual image reconstruction we aim to identify the most likely GMM component per patch, treating this as a discrete nuisance-parameter that is optimized along with the image $\mathbf{x}$.

@@

    Where $\mathcal{N}$ is the Gaussian normal distribution. If we knew the relevant component, $k_n$, for patch $\mathbf{x}_n$ we could derive the conditional log-prior distribution directly from Equation~\ref{eq:gmm}:
    %
    \begin{equation}
    \begin{split}
    \mathcal{P}_{\rm GMM}(\mathbf{x}_n|k_n) = -\frac{1}{2} \left[ \right. d \log(2\pi)
    + \log(|\boldsymbol \Sigma_{k_n}|)\\
    + (\mathbf{x}_n - \boldsymbol{\mu}_{k_n})^T \Sigma^{-1}_{k}(\mathbf{x}_n - \boldsymbol{\mu}_{k_n}) \left. \right].
    \label{eq:cond.log.prior}
    \end{split}
    \end{equation}
    %
    If we split the image $\mathbf{x}$ into multiple non-overlapping patches, for each patch $\mathbf{x}_n$, we can find the GMM component $\hat{k}_n$, which maximises the conditional log-prior distribution given in Equation~\ref{eq:cond.log.prior}:
    %
    \begin{equation}
        \hat{k}_n = \underset{k \in \{1, ..., K\} }{\argmax}{\mathcal{P}_{\rm GMM}(\mathbf{x}_n|k)}
        \label{eq:khat}
    \end{equation}
    \vspace{0.2em}
    %
    As the number of components in the GMM is typically limited to $\mathcal{O}(10^2)$ we can optimize the conditional log-prior distribution by simply evaluating it for each of the GMM components and choose the one with the maximum value. Finally, to evaluate the (optimized) log-prior of the full image we sum the maximum values of the individual patches: 
    %
    \begin{equation}
    \label{eq:prior-total}
        \mathcal{P}(\mathbf{x}, \hat{\mathbf{k}}) = 
        %\mathcal{P}(\mathbf{x} | \hat{\mathbf{k}})  + \mathcal{P}(\hat{\mathbf{k}}) =
        \sum_{n = 1}^N \mathcal{P}_{\rm GMM}(\mathbf{P}_n \mathbf{x} | \hat{k}_n) +
        \sum_{n = 1}^N \log \pi_{\hat{k}_n},
    \end{equation}   
    %
    where $\mathbf{P}_n$ is again the matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed.

@@

    Apart from this, the process for choosing the most likely GMM prior component remains unmodified. However, we note that we neglected the prior dependence between neighbouring patches\footnote{That is, we continue to evaluate the log-prior distribution of the full image by summing over the overlapping patches as in Equation~\ref{eq:prior-total}, even though overlapping patches include common pixels and are thus clearly not \aprioir independent.}. As the choice of the GMM component per patch is discrete, a simultaneous gradient based optimization of all patches is not easily possible, because we cannot compute gradients across GMM components. This would require a continuous representation or interpolation between GMM components.

    \subsubsection{Image Dynamic Range and Patch Normalization}
    \label{ssec:patch-norm}
    Astronomical images typically show a much higher dynamic range compared to natural images. To normalize the patches \cite{Zoran2011} proposed to subtract the patch mean before learning the GMM. This leads to the GMM modeling the same morphological features at different levels of contrast as different components. While this proved to be beneficial in the context of reconstructing natural images it does not scale to images with high dynamic range, because every morphological feature would need to be present at many levels of intensity. While this could be handled by introducing significantly more components to the GMM, it seems unfeasible in practice due to the limited availability of training data and computational resources. For this reason we instead propose a modified normalization for the patches:
    %
    \begin{equation}
        \label{eq:patch-norm}
        \tilde{\mathbf{x}}_n = \frac{\mathbf{x}_n - \Bar{\mathbf{x}}_n}{\Bar{\mathbf{x}}_n}
    \end{equation}
    %
    By dividing by the mean of the patch, $\Bar{\mathbf{x}}_n$, the structure in the patch becomes independent of the total brightness the patch contains. This results in patches being grouped only by their pixel to pixel contrast and not by their overall brightness. This corresponds to the assumption that the structure in the patch is \aprioir independent of the total flux contained in the patch, which we consider a valid assumption for astronomical images. This also generalizes to the prior distribution of the whole image described by Equation~\ref{eq:prior-total}. The value of the prior distribution is independent of the total flux of the image. The prior distribution simply regularizes the structure in the image, while the overall intensities are typically well constrained by data as quantified by the Poisson likelihood term. Thus we simply assume a uniform prior distribution for the patch intensities. Algorithmically, we replace $\mathbf{x}_n$ with $\tilde{\mathbf{x}}_n$ when evaluating Equations~\ref{eq:gmm} and \ref{eq:khat} and transform $\mathbf{P}_n\mathbf{x}$ via Equation~\ref{eq:patch-norm} when evaluating Equation~\ref{eq:prior-total}.

@@

    \subsection{Systematic Errors of Predicted Counts}
    Each observation $\mathbf{D}_j$ is subject to systematic errors. For example, the pointing accuracy of the telescope when repeatedly pointed to the same object or when using different types of telescopes for observation, the reconstructed position of an astronomical object may vary among the observations. The process of correcting for these inaccuracies is called \textit{astrometry} and is a standard procedure in for example \chandra data analysis. A second example of systematic errors is the variation of background emission between multiple observations.

    To account for both systematic effects we extend the model in Equation~\ref{eq:model} for the predicted counts, by a background normalization factor $\alpha_j$ and a linear shift in position $\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})$, which is specific for each observation. This yields the following modified model:
    %
    \begin{equation}
        \label{eq:model-npred-calibration}
        \mathbf{\lambda}_j = \mathbf{E}_j \cdot  \left(\mathrm{PSF}_j \circledast \phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})\right) + \alpha_j \cdot \mathbf{B}_j
    \end{equation}
    %
    In practice one observation is used as a reference and its corresponding systematic shifts in position $\delta_{x,j}, \delta_{y,j}$ are frozen during optimization. An example of the effect of the positional shift and background normalization can be seen in Figure~\ref{fig:example-fermi-lat-non-calibrated}.

@@

    We also experimented with computing the diagonal of the Hessian of the posterior distribution evaluated at the MAP estimate. However we found this numerically challenging because of the random cycle spinning.

@@

    We implemented the \jolideco method as an independent Python package, based on PyTorch as the computational back-end.  We defined a modular, object oriented code structure, to allow parts of the algorithm to be flexible and interchangeable, such as the choice of the patch normalization, the choice of GMM, optimization methods, and serialisation formats for the reconstruction results as well as the corresponding diagnostic information, such as the trace of the posterior and prior likelihood values. This will simplify future improvement and extension of the package as well as the adoption of \jolideco by the astronomical community. The package is available on the platform GitHub\footnote{\url{https://github.com/jolideco/jolideco}}.

@@

    \subsection{Learning GMMs}
    \label{ssec:jolideco-gmm-library}
    The patch-based image prior described in Section~\ref{sec:patch-prior} requires a pre-trained GMM. We train the GMM by choosing a set of reference images, splitting them into patches of size $8\times8$ pixels and learning a GMM with $K$ components from the whole distribution of patches. In general the strength of the GMM prior is its high \enquote{expressiveness}. This means the prior distribution contains a large variety of structural shapes and has freedom to adapt to the data in the optimization process by choosing the most likely GMM component for each patch in the image. With radio data, \cite{Bouman2016} found their results to be mostly independent of the choice of data used for learning the GMM prior.

@@

    The ground truth and corresponding expected counts for several exemplar combinations of source and instrument scenarios are shown in Figure~\ref{fig:scenarios}.
